Measurement of ($\pi^-$-Ar) and ($K^+$-Ar) total hadronic cross sections at the LArIAT experiment

### 2018-05-31 Chapter-4/4-Methodology.tex
s

\subsection{TPC Selection: Shower Removal}\label{ch:electrons}
In the case of the ($\pi^-$,Ar) cross section, the resolution of  beamline mass spectrometer is not sufficient to select a beam of pure pions. In fact,  muons which are close in mass to the pions and relativistic electrons survive the selection on the beamline mass.  It is important to notice that the composition of the negative polarity beam is mostly pions, as will be discussed in section \ref{ch:beamlineComposition}.
Still, we devise a selection on the TPC information to mitigate the presence of electrons in the sample used for the pion cross section. The selection relies on the different topologies of a pion and an electron event when propagating in liquid argon: while the former will trace a track inside the TPC active volume, the latter will tend to ``shower", i.e. interact with the medium, producing bremsstrahlung photons which pair convert into several short tracks. In order to remove the shower topology, we create a region of interest (ROI) around the TPC track corresponding to the beamline particle. We look for short tracks contained in the ROI, as depicted in Figure \ref{fig:showerFilt}:  if more then 5 tracks shorter than 10 cm are in the ROI, we reject the event. Line 7 in  Table \ref{tab:beamlineDataSelection} shows the number of events surviving this selection; that table also shows that this selection is applied after the wire chamber to TPC match (discussed in the next section). The WC2TPC match already lower the presence of electrons in the sample, which is further reduced by the shower filter.

@@

\begin{equation}
 \sigma^{\pi^-}_{TOT}(E_{i})  = \frac{1}{n \delta X}\frac{N^{\pi^-}_{ \text{Int}} (E_{i})}{N^{\pi^-}_{ \text{Inc}}(E_{i})}.
\label{eq:thinTargetXSSolved2}
\end{equation}

Unfortunately, this is not the case. In fact, the selection used to isolate pions in the LArIAT beam allows for the presence of some muons and electrons as background, while the kaon selection allows for a small contamination of protons (see Section \ref{ch:beamlineComposition}). Also, the LArIAT TPC tracking algorithm is not 100\% efficient in determining the interaction point. Therefore we need to apply two corrections evaluated on the MC in order to extract the final cross section from LArIAT data: i) a background subtraction and ii) a correction for reconstruction effects. 
Still using the pion case as example, we estimate the pion cross section in each energy bin changing  Equation \ref{eq:thinTargetXSSolved2} into
\begin{equation}
 \sigma^{\pi^-}_{TOT}(E_{i})  =\frac{1}{n\text{ } \delta X}\frac{N^{\pi^-}_{ \text{Int}} (E_{i})}{N^{\pi^-}_{ \text{Inc}}(E_{i})} = \frac{1}{n \text{ }\delta X}\frac{ \epsilon^{\text{Inc}}(E_i) [ N^{ \text{TOT}}_{ \text{Int}} (E_{i}) - B_{ \text{Int}} (E_i)] }{   \epsilon^{\text{Int}}(E_i) [N^{ \text{TOT}}_{ \text{Inc}}(E_{i}) - B_{ \text{Inc}} (E_i)]},
\label{eq:True}
\end{equation}

>>>>

\subsection{TPC Selection: Shower Removal}\label{ch:electrons}
In the case of the ($\pi^-$,Ar) cross section, the resolution of  beamline mass spectrometer is not sufficient to select a beam of pure pions. In fact,  muons which are close in mass to the pions and relativistic electrons survive the selection on the beamline mass.  It is important to notice that the composition of the negative polarity beam is mostly pions, as will be discussed in section \ref{ch:beamlineComposition}.
Still, we devise a selection on the TPC information to mitigate the presence of electrons in the sample used for the pion cross section. The selection relies on the different topologies of a pion and an electron event when propagating in liquid argon: while the former will trace a track inside the TPC active volume, the latter will tend to ``shower", i.e. interact with the medium, producing bremsstrahlung photons which pair convert into several short tracks. In order to remove the shower topology, we create a region of interest (ROI) around the TPC track corresponding to the beamline particle. We look for short tracks contained in the ROI, as depicted in Figure \ref{fig:showerFilt}:  if more then 5 tracks shorter than 10 cm are in the ROI, we reject the event. Line 7 in  Table \ref{tab:beamlineDataSelection} shows the number of events surviving this selection; that table also shows that this selection is applied after the beamline event is matched to TPC particle (discussed in the next section). This match already lowers the presence of electrons in the sample, which is further reduced by the shower filter.

@@

\begin{equation}
 \sigma^{\pi^-}_{TOT}(E_{i})  = \frac{1}{n\text{ } \delta X}\frac{N^{\pi^-}_{ \text{Int}} (E_{i})}{N^{\pi^-}_{ \text{Inc}}(E_{i})}.
\label{eq:thinTargetXSSolved2}
\end{equation}

Unfortunately, this is not the case. In fact, the selection used to isolate pions in the LArIAT beam allows for the presence of some muons and electrons as background, while the kaon selection allows for a small contamination of protons (see Section \ref{ch:beamlineComposition}). Also, the LArTPC tracking algorithm is not 100\% efficient in determining the interaction point. This inefficiency occurs in two fashions: i) the tracking algorithm does not stop at the interaction point and continues adding hits from a particle past it (this happens especially in the case of shallow elastic scattering), ii) the tracking stops prematurely. These two cases have different consequences on the population of the interacting and incident distributions. In the first case, the interacting histogram will be underpopulated and the incident histogram might be overpopulated. In case of premature end of tracking, the interacting histogram will be overpopulated at energies greater than the eventual interaction, while the incident histogram will be underpopulated. Given the importance of tracking for the cross section measurements,  we report an optimization to maximize the identification of the interaction point in Appendix \ref{ch:AppendixTrack}.

Therefore, we apply two corrections evaluated on MC in order to extract the final cross section from LArIAT data: i) a background subtraction and ii) a correction for reconstruction effects. 
Still using the pion case as example, we estimate the pion cross section in each energy bin changing  Equation \ref{eq:thinTargetXSSolved2} into
\begin{equation}
 \sigma^{\pi^-}_{TOT}(E_{i})  =\frac{1}{n\text{ } \delta X}\frac{N^{\pi^-}_{ \text{Int}} (E_{i})}{N^{\pi^-}_{ \text{Inc}}(E_{i})} = \frac{1}{n \text{ }\delta X}\frac{ \epsilon^{\text{Inc}}(E_i) [ N^{ \text{TOT}}_{ \text{Int}} (E_{i}) - B_{ \text{Int}} (E_i)] }{   \epsilon^{\text{Int}}(E_i) [N^{ \text{TOT}}_{ \text{Inc}}(E_{i}) - B_{ \text{Inc}} (E_i)]},
\label{eq:True}
\end{equation}

### 2018-05-31 Chapter-5/5-Preparatory.tex
s

\section{Tracking Studies}\label{sec:TrackingStudies}
The tracking of hadrons in the TPC determines both the beamline to TPC handshake and the identification of the interaction point within the TPC. Thus, it plays a fundamental role in the cross section measurements. We performed several studies geared towards the optimization of the LArSoft software package for tracking in the TPC. In particular, we studied a suitable set of parameters for the WC2TPC match and we optimized the clustering algorithm to maximize the efficiency of finding the interaction point on MC. Given the technical nature of these studies, we report them in Appendix \ref{ch:AppendixTrack}. 
We only report here the evaluation of  the angular resolution of the tracking algorithm in data and MC, due to its important implications on the physics measurement.

@@

>>>>

\section{Tracking Studies}\label{sec:TrackingStudies}
The tracking of hadrons in the TPC determines both the beamline to TPC handshake and the identification of the interaction point within the TPC. Thus, it plays a fundamental role in the cross section measurements. We performed several studies geared towards the optimization of the LArSoft software package for tracking in the TPC. In particular, we studied a suitable set of parameters for the WC2TPC match and we optimized the clustering algorithm to maximize the efficiency of finding the interaction point on MC. Given the technical nature of these studies, we report them in Appendix \ref{ch:AppendixTrack}. 
We only report here the evaluation of  the angular resolution of the tracking algorithm and a study of the track pitch in data and MC, due to their important implications on the physics measurement.

@@

\subsection{Track Pitch}\label{sec:trackPitch}
The granularity of the LArTPC detector leads to the fine-sampling of a hadron inside the argon active volume. This detector granularity translates into the ``track pitch" quantity, which is the distance between two consecutive 3D points of the tracking. The track pitch determines the width of the slab $\delta X$ in the cross section Equation \ref{eq:thinTargetXSSolved2}. Figure \ref{fig:Pitch} shows the pitch distribution in data (black) and MC (red) for the pion combined sample on the left and the kaon sample on the right. The MC and data distributions agree rather well.   Thus, we estimate the slab width to be $\delta X = 0.47 \pm 0.03 $ cm.

\begin{figure}[p]
\centering
\includegraphics[width=0.48\textwidth]{Chapter-5/Images/PitchPi.pdf}
\includegraphics[width=0.48\textwidth]{Chapter-5/Images/PitchK.pdf}
\caption{\emph{Left:} Pitch distribution for the pion combined sample 60A and 100A, data displayed in black, MC in red. 
\emph{Right:} Pitch distribution for the kaon sample, data displayed in black, MC in red.}
\label{fig:Pitch}
\end{figure}

### 2018-05-31 Chapter-6/6-PionCrossSection.tex
s

$N^{\text{TOT}}_{\text{Inc}}$ and $N^{\text{TOT}}_{\text{Int}}$ are not independent. In fact, the population of a given bin for the interacting histogram always implies at least the population of the same bin in the incident histogram  (and possibly other incidents bins at higher energies). Thus, we conservatively calculate the statistical uncertainty on the cross section as 
\begin{equation}
\delta\sigma_{TOT}(E) = \sigma_{TOT}(E) \Big(\frac{\delta N^{\text{TOT}}_{\text{Int}}}{N^{\text{TOT}}_{\text{Int}}}+\frac{\delta N^{\text{TOT}}_{\text{Inc}}}{N^{\text{TOT}}_{\text{Inc}}}\Big) 
\end{equation}
where:
\begin{eqnarray}
\delta N^{\text{TOT}}_{\text{Inc}} = \sqrt[]{N^{\text{TOT}}_{\text{Inc}}} \\
\delta N^{\text{TOT}}_{\text{Int}} = \sqrt[]{N^{\text{TOT}}_{\text{Int}}\Big(1-\frac{ N^{\text{TOT}}_{\text{Int}}}{N^{\text{TOT}}_{\text{Inc}}}\Big)}.
\end{eqnarray}

\subsection{Treatment of Systematics} \label{ch:SysUncertaintyXSRaw}
The only systematic effect considered in the measurement of the raw cross section results from the propagation of the uncertainty associate with the measurement of the kinetic energy at each argon slab. As shown in Section \ref{ch:kinEn}, the uncertainty on the kinetic energy of a pion candidate at the j$^{th}$ slab of argon  is given by

@@

\subsection{Correction for Reconstruction Effects}\label{ch:EFFXS}
The interaction point for a particle in the selected sample for the total hadronic cross section analysis is the last point of its track that lies inside the LArTPC fiducial volume. This definition holds regardless the type of the interaction, i.e. if the TPC track ends within the fiducial volume, its last point will be the interaction point, no matter what the products of the interaction look like; on the other hand, if the track crosses the boundaries of the fiducial volume, the particle will be considered ``through going" and no interaction point will be found.  Given this definition, it is evident that we rely on the tracking algorithm to discern where the interaction occurred in the TPC  and correctly end the tracking. The tracking algorithm has an intrinsic angle resolution as shown in Section \ref{sec:angleRes}, which limits its efficiency, especially in the case of elastic scattering occurring a low angles. \textcolor{red}{Plus, there are instance where INSERT HERE THE STUFF ABOUT THE MIGRATION}
Thus, we need to apply a correction accounting for mis-reconstruction of the interaction point in order to retrieve the true cross section.  This correction is evaluated separately for the interacting and incident histograms bin by bin, namely $\epsilon^{\text{int}}(E_i)$ and  $\epsilon^{\text{inc}}(E_i)$, and applied in the cross section formula as shown in  Equation \ref{eq:C}.

@@

In Section \ref{sec:angleRes}, we estimated the angular resolution for data and MC to be $\bar\alpha_{Data} = (5.0 \pm 4.5) \text{ deg}$  and 
$\bar\alpha_{MC} = (4.5 \pm 3.9) \text{ deg}$, respectively.  Most interaction angles smaller than the angular resolution will thus be indistinguishable  for the reconstruction. Thus, we claim we are able to  measure the cross section for interaction angles greater than 5.0 deg. Geant4 simulates interactions at all angles, as shown in figure \ref{fig:trueScatteringAngle}. In order to calculate the correction for reconstruction effects,  we select events which have an interaction angle greater than a given $\alpha_{res}$ to construct the true interacting and incident histograms (the denominator of the correction).

@@

 The systematics on this correction is estimated by varying the value of $\alpha_{res}$ between 0 deg and 4.5 deg and propagating the uncertainty on the cross section.

@@

\clearpage
\section{Results}\label{ch:FinalPion}
Figure \ref{fig:FinalXSPion} shows the measurement of the ($\pi^-$-Ar) total hadronic cross section for  scattering angles greater than 5$^\circ$, as the result of the background subtraction and reconstruction effects correction to the raw cross section. The top left plot is the measurement obtained on the 60A data, statistical and systematic uncertainty shown in red. The top right plot is the measurement obtained on the 100A data, statistical uncertainty and systematic uncertainty shown in blue. The bottom plot shows the two measurements overlaid. In all three plot, the Geant4 prediction for the total hadronic cross section for angle scattering greater than 5$^\circ$ is displayed in green.

The systematic uncertainty on the cross section is the sum in quadrature of the statistical uncertainty, the systematic uncertainty related to the kinetic energy measurement, the systematic uncertainty related to the beam composition and the systematic uncertainty related to the reconstruction effects correction.

>>>>

$N^{\text{TOT}}_{\text{Inc}}$ and $N^{\text{TOT}}_{\text{Int}}$ are not independent. In fact, the population of a given bin for the interacting histogram always implies at least the population of the same bin in the incident histogram  (and possibly other incidents bins at higher energies). Thus, we conservatively calculate the statistical uncertainty on the cross section as 
\begin{equation}
\frac{\delta\sigma_{TOT}(E)}{\sigma_{TOT}(E)} =  \Big(\frac{\delta N^{\text{TOT}}_{\text{Int}}}{N^{\text{TOT}}_{\text{Int}}}+\frac{\delta N^{\text{TOT}}_{\text{Inc}}}{N^{\text{TOT}}_{\text{Inc}}}\Big) 
\end{equation}
where:
\begin{eqnarray}
\delta N^{\text{TOT}}_{\text{Inc}} = \sqrt[]{N^{\text{TOT}}_{\text{Inc}}} \\
\delta N^{\text{TOT}}_{\text{Int}} = \sqrt[]{N^{\text{TOT}}_{\text{Int}}\Big(1-\frac{ N^{\text{TOT}}_{\text{Int}}}{N^{\text{TOT}}_{\text{Inc}}}\Big)}.
\end{eqnarray}

\subsection{Treatment of Systematics} \label{ch:SysUncertaintyXSRaw}
The systematic effects considered in the measurement of the raw cross section results from the uncertainty on the track pitch  ($\delta X = 0.47 \pm 0.03 $ cm, shown in \ref{sec:trackPitch})  and from the propagation of the uncertainty associate with the measured kinetic energy at each argon slab. As shown in Section \ref{ch:kinEn}, the uncertainty on the kinetic energy of a pion candidate at the j$^{th}$ slab of argon  is given by

@@

\subsection{Correction for Reconstruction Effects}\label{ch:EFFXS}
The interaction point for a particle in the selected sample for the total hadronic cross section analysis is the last point of its track that lies inside the LArTPC fiducial volume. This definition holds regardless the type of the interaction, i.e. if the TPC track ends within the fiducial volume, its last point will be the interaction point, no matter what the products of the interaction look like; on the other hand, if the track crosses the boundaries of the fiducial volume, the particle will be considered ``through going" and no interaction point will be found.  Given this definition, it is evident that we rely on the tracking algorithm to discern where the interaction occurred in the TPC  and correctly end the tracking. The tracking algorithm has an intrinsic angle resolution as shown in Section \ref{sec:angleRes}, which limits its efficiency, especially in the case of elastic scattering occurring a low angles. Also, as mentioned in Section \ref{ch:MCCorrections}, the tracking algorithm can prematurely stop, mis-identifying an interaction point. 
Thus, we need to apply a correction accounting for mis-reconstruction of the interaction point in order to retrieve the final cross section.  This correction is evaluated separately for the interacting and incident histograms bin by bin, namely $\epsilon^{\text{int}}(E_i)$ and  $\epsilon^{\text{inc}}(E_i)$, and applied in the cross section formula as shown in  Equation \ref{eq:C}.

@@

In Section \ref{sec:angleRes}, we estimated the angular resolution for data and MC to be $\bar\alpha_{Data} = (5.0 \pm 4.5) \text{ deg}$  and 
$\bar\alpha_{MC} = (4.5 \pm 3.9) \text{ deg}$, respectively.  Most interaction angles smaller than the angular resolution will thus be indistinguishable  for the reconstruction. Thus, we claim we are able to  measure the cross section for interaction angles greater than 5.0 deg. Geant4 simulates interactions at all angles, as shown in figure \ref{fig:trueScatteringAngle}. In order to calculate the correction for reconstruction effects,  we select events which have an interaction angle greater than a given $\alpha_{res}$ to construct the true MC interacting and incident histograms (the denominator of the correction).

@@

 The systematics on this correction is estimated by varying the value of $\alpha_{res}$ between 0 deg and 5.0 deg and propagating the uncertainty on the cross section.

@@

\clearpage
\section{Results}\label{ch:FinalPion}
Figure \ref{fig:FinalXSPion} shows the measurement of the ($\pi^-$-Ar) total hadronic cross section,

\begin{equation}
 \sigma^{\pi^-}_{TOT}(E_{i})  =\frac{1}{n\text{ } \delta X}\frac{N^{\pi^-}_{ \text{Int}} (E_{i})}{N^{\pi^-}_{ \text{Inc}}(E_{i})} = \frac{1}{n \text{ }\delta X}\frac{ \epsilon^{\text{Inc}}(E_i) [ N^{ \text{TOT}}_{ \text{Int}} (E_{i}) - B_{ \text{Int}} (E_i)] }{   \epsilon^{\text{Int}}(E_i) [N^{ \text{TOT}}_{ \text{Inc}}(E_{i}) - B_{ \text{Inc}} (E_i)]},
\end{equation}

for  scattering angles greater than 5$^\circ$, as the result of the background subtraction and reconstruction effects correction to the raw cross section. The top left plot is the measurement obtained on the 60A data, statistical and systematic uncertainty shown in red. The top right plot is the measurement obtained on the 100A data, statistical uncertainty and systematic uncertainty shown in blue. The bottom plot shows the two measurements overlaid. In all three plot, the Geant4 prediction for the total hadronic cross section for angle scattering greater than 5$^\circ$ is displayed in green.

The systematic uncertainty on the cross section is the sum in quadrature of the statistical uncertainty, the systematic uncertainty calculated for the raw cross section, the systematic uncertainty related to the beam composition and the systematic uncertainty related to the reconstruction effects correction.

### 2018-05-31 Chapter-7/7-KaonCrossSection.tex
s

where $N^{\text{TOT}}_{\text{Int}}$  is the measured number of particles interacting at kinetic energy $E_i$, $N^{\text{TOT}}_{\text{Inc}}$ is the  measured  number of particles incident  on an argon slice at  kinetic energy $E_i$,  $n$ is the density of the target centers  and $\delta X$ is the thickness of the argon slice. The density of the target centers and the slab thickness are $n = 0.021\cdot10^{24} \text{ cm}^{-3} $ and  $\delta X=0.47\text{ cm}$, respectively.

@@

As in the pion case, the only systematic effect considered in the measurement of the raw cross section results from the propagation of the uncertainty associate with the measurement of the kinetic energy at each argon slab. For kaons, the uncertainty on the kinetic energy of a candidate at the j$^{th}$ slab of argon  is given by

@@

\section{Results}\label{ch:FinalKaon}
Figure \ref{fig:FinalXSKaon} show the measurement of the ($K^+$-Ar) total hadronic cross section for  scattering angles greater than 5$^\circ$, as the result of the background subtraction and reconstruction  effects  correction to the raw cross section. The plot shows the measurement obtained on the full dataset, statistical and systematic uncertainty in black. The Geant4 prediction for the total hadronic cross section for angle scattering greater than 5$^\circ$ is displayed in green.

The systematic uncertainty on the cross section is the sum in quadrature of the statistical uncertainty, the systematic uncertainty related to the kinetic energy measurement and the systematic uncertainty related to the reconstruction  effects  correction.

>>>>

where $N^{\text{TOT}}_{\text{Int}}$  is the measured number of particles interacting at kinetic energy $E_i$, $N^{\text{TOT}}_{\text{Inc}}$ is the  measured  number of particles incident  on an argon slice at  kinetic energy $E_i$,  $n$ is the density of the target centers  and $\delta X$ is the thickness of the argon slice. The density of the target centers and the slab thickness are $n = 0.021\cdot10^{24} \text{ cm}^{-3} $ and  $\delta X=0.47 \pm 0.03\text{ cm}$, respectively.

@@

As in the pion case, the systematic effects considered in the measurement of the raw cross section results from the uncertainty on the pitch ($\delta X = 0.47 \pm 0.03$ cm) and from the propagation of the uncertainty associate with the measurement of the kinetic energy at each argon slab. For kaons, the uncertainty on the kinetic energy of a candidate at the j$^{th}$ slab of argon  is given by

@@

\section{Results}\label{ch:FinalKaon}
Figure \ref{fig:FinalXSKaon} show the measurement of the ($K^+$-Ar) total hadronic cross section,
\begin{equation}
 \sigma^{K^+}_{TOT}(E_{i})  =\frac{1}{n\text{ } \delta X}\frac{N^{K^+}_{ \text{Int}} (E_{i})}{N^{K^+}_{ \text{Inc}}(E_{i})} = \frac{1}{n \text{ }\delta X}\frac{ \epsilon^{\text{Inc}}(E_i) [ N^{ \text{TOT}}_{ \text{Int}} (E_{i}) - B_{ \text{Int}} (E_i)] }{   \epsilon^{\text{Int}}(E_i) [N^{ \text{TOT}}_{ \text{Inc}}(E_{i}) - B_{ \text{Inc}} (E_i)]},
\end{equation}

for  scattering angles greater than 5$^\circ$, as the result of the background subtraction and reconstruction  effects  correction to the raw cross section. The plot shows the measurement obtained on the full dataset, statistical and systematic uncertainty in black. The Geant4 prediction for the total hadronic cross section for angle scattering greater than 5$^\circ$ is displayed in green.

The systematic uncertainty on the cross section is the sum in quadrature of the statistical uncertainty, the systematic uncertainty calculated for the raw cross section and the systematic uncertainty related to the reconstruction  effects  correction.
